{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2023-09-06 Docs/manuscript.Rnw
Edits to current draft, esp. discussion

% To optimize ecosystem services for habitat conservation and understand how plants cope with environmental variations, predicting the biological impacts of climate change is of critical importance \citep{botero15}. One of the most important parts of accurate prediction is understanding the relative importance of local adaptation versus plasticity \citep{chevin10}. 
Predicting the biological impacts of climate change has made understanding how organisms cope with environmental variation more urgent \citep{botero15}. In particular, the relative importance of plasticity versus genetic adaptation is vital for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptation \citep{chevin102, snell18}.

@@

% emwJun4: I really love the info you added here, but we have to be careful to not focus too deeply on synchrony as it is a very specific topic and our work is broader than that (sorry if I confused you by my request for 'refs' -- I mostly wanted just citations; I suspect we can use some of the text for the discussion so it will not go to waste). We also need to keep paragraphs themselves and our intro overall short, so I cut here. Thus I slimmed this way down and kept many references, but less of the examples.
Many of the currently observed responses to climate change appear to be mostly plastic \citep{burton22,zettle21,bonamour19, king17}, including the most reported biological response to climate change---shifting phenology. Phenology (the timing of recurring seasonal events) governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and exploit the resources of each growing season \citep{chuine10,hanninen11,rytteri21,posle18}. As such, phenology plays a significant role in determining fitness for both plants \citep{guo22,chuine01} and animals \citep{wann19,renner18,chu17}.

Shifted phenology in recent decades---with many events moving several days per decade \citep{vita21,khar18,Menzel06}---has led to concerns about fitness consequences, and the limits of possible future shifts. Future phenological shifts will depend on how much phenology is determined by plasticity versus adaptation; yet our understanding of the balance of these two approaches to variation is limited, even for species groups that are critical to both forecasting and have been well studied, such as trees.

@@

Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies  (i.e., studies---often for forestry purposes---where researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype) \citep{AitkenBemmels16, Alberto13}. This work has established common clines in fall phenology, as source locations with shorter growing season (poleward and higher elevations) exhibit earlier stopping of growth (e.g., budset). Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold that trigger budset), driven by local adaptation to the local growing season \citep{Alberto13,Savolainen07}. In contrast, spring phenology appears far more plastic \citep{AitkenBemmels16} and is determined more strongly by temperature \citep{flynn18}. Many studies, however, have argued that spring phenology shows high enough levels of adaptation to matter biologically.  % alinaJuly20: im not sure what important levels of adaptation means here, %emwaug1: I meant that the 'amount' of local adaptation is high enough to matter (contrasting with the reality that if you have a big enough sample you can often find a small difference that is 'significant' but it might not matter biologically).

Results, however, have not always been consistent. Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (for example, \emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}). This has led to some debate over the prevalence and importance of adaptation in spring tree phenology. While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they could vary by continent.

Continental differences in patterns of adaptation versus plasticity could be driven by climatic differences, as North American springs are more variable across years than European ones \citep{tward21,zohner2017,schwartz00}. Such high temporal variability means that distant sites can effectively experience the same spring climate, but in different years. Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{zohner2017}, but are poorly controlled compared to traditional provenance trials, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}. Even for more carefully designed common garden differences in study design---especially studied species---may complicate understanding what underlies potential trends across continents, with European studies focusing more on deciduous angiosperms and North American studies focusing often on evergreen gymnosperms.

@@

\section{Results}
Our final dataset included seven angiosperm and eight gymnosperm species from 17 studies, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019. Seven species also had fall event information available. 
<<label=source_spring_models, echo=FALSE>>=
#load models and data to report numbers of things
# "C:/Users/alina/Documents/git/localadaptclim"
# setwd("C:/Users/alina/Documents/git/localadaptclim/Docs")
#getwd()
source("../Analyses/source_spring_models_for_sweave.R")
@

@@

In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}]) and the decrease in MAT (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}]), meaning fall events were earlier where provenance MAT was lower (Fig. \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ). This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig. \ref{figure:continental_spptype_effect}A).

@@

While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts local adaptation. For spring events, we considered overlap across March to May. However, results were not qualitatively different than using MAT (See Fig. \ref{supplement-figure:overlap_scatterplot} in Supplements). We observed very weak effects of climate overlap on spring events (\Sexpr{format(round(abs(mean(all_species_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent increase in climate overlap), nearly identical across angiosperms (\Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)}]) and gymnosperms (\Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}]). Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).

@@

Overall, our results demonstrated inconsistent and weak clines in spring events across North America and Europe. In contrast, fall events generally showed much stronger clines, especially in North America, and in support of many previous studies \citep{AitkenBemmels16, Alberto13}. While previous studies have suggested spring events are far more plastic compared to fall events \citep{Li97,farmer93,mimura07}, our study provides the first major test of this across continents and species and suggests overall no general trend for clines in spring phenology. While not a direct test, our results suggest that spring events are highly plastic; they thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures, in line with most current climate change impact studies \citep{IPCC22}

Trends between spring phenology and latitude or MAT were weak, but suggested the possibility of diverging results that could mute an overall trend (albeit a much weaker one than for fall phenology). Clines with MAT diverged between angiosperms and gymnosperms. Combined with our finding of stronger clines for fall phenology in North America, our results suggest that data on more species tested across continents could yield important insights into plasticity versus adaptation limits for spring and fall phenology, which would greatly aid forecasting.

We found the coarse metrics of provenance latitude and MAT were generally good predictors of phenology, performing better than climate overlap. Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in months close to the events, at least in the narrow range of northern temperature and boreal forests we studied.

\subsection{Variation across continents and species groupings}

Our results indicated that the effects of provenance latitude and MAT on fall events were similar across angiosperm and gymnosperm species: Fall events delay as MAT increases for both leaf types, slightly more so for gymnosperms, explained by sensitivity to day-length signals. However, the two leaf types exhibited slightly diverging trends in spring events:  advancing in angiosperms and delaying in most gymnosperms as MAT increases. Such differences make sense as timing of budburst and leafout varies by leaf type. Gymnosperms photosynthesize without leafout and thus generally leafout later than angiosperms, after much risk of spring frost \citep{panchen14}. Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring. According to \citet{salk20}, species with narrow xylem such as coniferous gymnosperms and diffuse-porous angiosperms did not exhibit the consistent phenological trends shown in ring-porous species. Therefore, there might be additional differences if we have more species and more information on leafout timing and frost risk approach.

While we noticed more prominent clines in fall events in North America, there was no strong evidence of overall continental differences in spring events, which might again suggest a higher degree of plasticity in spring events. Meanwhile, most North American gardens and provenances included in this study were limited to the Pacific Northwest region, which might mask the kind of climate pattern we were hoping to see from the entirety of North America (Fig. \ref{figure:map_gardens}).

\subsection{Adaptation in spring vs fall phenology: Implications for climate change responses}

Our results suggest that current advances in spring phenology may continue, but delays in fall phenology could depend on the speed of adaptation. While some previous chamber studies have found potential adaptation, these could have been influenced by chilling at the provenance site \citep{ettinger20}. However, field studies and common garden experiences do not have this effect. Spring events from the studies we looked at appeared to exhibit high plasticity, while fall events showed strong clines, which is consistent with previous work \citep{AitkenBemmels16, Alberto13}. These findings have important implications for growing discussion of how shifting phenology and tree growth relate as our results suggest there are strong limits on shifts in fall phenology \citep{zohner23, zani20}.

It is worth noting that these trends likely vary by species and depend on frost tolerance strategy. If current advances in spring phenology continues, it could suggest shifts in spring frost. In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19}. Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}. Further, plants that budburst earlier can take advantage of a longer growing season and avoid competition \citep{guo22}. However, differences across species means that this will not impact all species the same. While fall event always exhibited co-gradient relationships (i.e. events advanced as we moved northward or up in elevation), spring events were more varied among species. The implication of such variations could be that spring leafout sets up flowering, and thus, some counter-gradient could better aid gene glow over larger geographical regions.

All of above are critical to predicting not just phenological change, but range expansions with climate change: An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction. Implications of small differences in spring phenology could mean different levels of gene flow.

Our observations reveal that although temperature by itself could serve as a reasonable indicator for predicting when fall events occur across various locations, it's crucial to acknowledge that the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}. High-latitude sites, for example, are likely restricted by the impact of photoperiod \citep{Gill15}. Diverse mechanisms that govern the timing of leaf senescence at the ecosystem level hold the potential to exert significant influence over not only the interactions occurring within plant communities but also the intricate process of carbon storage within ecosystems. With the anticipated escalation of global temperatures in the forthcoming century, these mechanisms are poised to play a pivotal role in shaping the dynamics of plant communities and the overall carbon balance of ecosystems.

>>>>

% To optimize ecosystem services for habitat conservation and understand how plants cope with environmental variations, predicting the biological impacts of climate change is of critical importance \citep{botero15}. One of the most important parts of accurate prediction is understanding the relative importance of local adaptation versus plasticity \citep{chevin10}. 
Predicting the biological impacts of climate change has made understanding how organisms cope with environmental variation more urgent \citep{botero15}. In particular, the relative importance of plasticity versus genetic adaptation is vital for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptation, but possibly limiting responses after the limits of plasticity are reached \citep{chevin102,snell18}.

@@

% emwJun4: I really love the info you added here, but we have to be careful to not focus too deeply on synchrony as it is a very specific topic and our work is broader than that (sorry if I confused you by my request for 'refs' -- I mostly wanted just citations; I suspect we can use some of the text for the discussion so it will not go to waste). We also need to keep paragraphs themselves and our intro overall short, so I cut here. Thus I slimmed this way down and kept many references, but less of the examples.
Many of the currently observed responses to climate change appear to be mainly plastic \citep{burton22,zettle21,bonamour19, king17}, including the most reported biological response to climate change---shifting phenology. Phenology---the timing of recurring seasonal events---governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and exploit the resources of each growing season \citep{chuine10,hanninen11,rytteri21,posle18}. As such, phenology plays a significant role in determining fitness for both plants \citep{guo22,chuine01} and animals \citep{wann19,renner18,chu17}.

Shifted phenology in recent decades---with many events moving several days per decade \citep{vita21,khar18,Menzel06}---has led to concerns about fitness consequences, and the limits of possible future shifts. While future phenological shifts will depend on how much phenology is determined by plasticity versus adaptation, our understanding of the balance of these two approaches to variation is limited. This is the case even for species groups that are critical to both forecasting and have been well studied, such as trees.

@@

Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies. In these studies---conducted often for forestry purposes---researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}. Such work has established common clines in fall phenology, as source locations with shorter growing seasons (poleward and higher elevations) exhibit earlier stopping of growth (such as budset). Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold required to trigger budset), driven by adaptation to the local growing season \citep{Alberto13,Savolainen07}. In contrast, spring phenology appears more plastic \citep{AitkenBemmels16} and is determined more strongly by temperature \citep{flynn18}. Many studies, however, have argued that spring phenology shows levels of adaptation that may be critical to forecasting and mitigation \citep{vitasse2009,Basler:2012}.

These contrasting studies highlight how inconsistent evidence for adaptation in tree spring phenology has been. Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have failed to find similar trends along latitudinal gradients (for example, \emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}). This has led to debate over the prevalence and importance of adaptation in spring tree phenology. While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they may be more studies and debated in Europe \citep{deans96,vitasse2009,Basler:2012}, raising the possibility that they could vary by continent. %EMWSep4: Check deans96 is accurate ref I added. Other refs I added to 'more' bib in case you do not have them.

Continental differences in patterns of adaptation versus plasticity could be driven by climatic differences, especially as North American springs are more variable across years than European ones \citep{tward21,zohner2017,schwartz00}. Such high temporal variability means that distant sites can effectively experience the same spring climate, but in different years. Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{zohner2017}, but are poorly controlled compared to traditional common garden studies, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}. Even for more carefully designed common gardens, differences in species studied or other differences in design may complicate understanding what underlies potential trends across continents.  %EMWSep4: Moved last point to discussion.

@@

\section{Results}
Our final dataset included seven deciduous angiosperm and eight evergreen gymnosperm species from 17 studies and 19 gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019. Seven species (five in North America and two in Europe) also had fall event information available. %EMWSep4 -- check my edits here; also if we don't have it somewhere in the results yet, please add gymnosperms/angiosperm species # in North America versus Europe here, something like: Most species in North American gardens were gymosperms (7/11 species) while most species were angiosperms in Europe (3/4 species).  
<<label=source_spring_models, echo=FALSE>>=
#load models and data to report numbers of things
# "C:/Users/alina/Documents/git/localadaptclim"
# setwd("C:/Users/alina/Documents/git/localadaptclim/Docs")
#getwd()
source("../Analyses/source_spring_models_for_sweave.R")
@

@@

%EMWSep4 -- small edits here, hopefully I did not screw up the formatting
In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more norther, cooler MAT sites (that is, they advanced strongly with provenance latitude: \Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}], and with decreasing MAT:\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}], Fig. \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ). This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig. \ref{figure:continental_spptype_effect}A).

@@

While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts provenance effects. For spring events, we considered overlap across March to May. However, results were not qualitatively different than using MAT (See Fig. \ref{supplement-figure:overlap_scatterplot} in Supplements). We observed very weak effects of climate overlap on spring events (\Sexpr{format(round(abs(mean(all_species_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent increase in climate overlap), nearly identical across angiosperms (\Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)}]) and gymnosperms (\Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}]). Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).

@@

Overall, our results demonstrated inconsistent and weak clines in spring events across North America and Europe. In contrast, fall events generally showed much stronger clines, especially in North America, and in support of many previous studies \citep{AitkenBemmels16, Alberto13}. While previous studies have suggested spring events are far more plastic compared to fall events \citep{Li97,farmer93,mimura07}, our study provides the first major test of this across continents and species and suggests no general trend for clines in spring phenology. Our results thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures \citep{IPCC22}

Trends between spring phenology and latitude or MAT were weak, but suggested the possibility of diverging results that could mute an overall trend---albeit a much weaker one than for fall phenology. We found angiosperm versus gymnosperm species diverged in their clines with MAT. Combined with our finding of much stronger clines for fall phenology in North America, these results support the idea of potential variation across continents and/or species type that may underlie the debate in whether spring events show important clinal variation. As clines with spring events were very weak, however, and gardens almost always focus on only one species, understanding these diverging results well enough to aid forecasting would take significant additional investment in common garden studies. %EMWSep4 -- we need to be careful in how many times we say 'we need more data' -- it's a common need, so it's better to focus on specifics of what the data suggests, or doesn't and say we need more data -- at most once in a paper. More on this below.

We found the coarse metrics of provenance latitude and MAT were generally good predictors of phenology, performing better than our more complex and data-rich metric of climate overlap. Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in  the temperate and boreal forest species we studied, yielding similar results metrics calculated specifically in the spring with daily climate data. %EMWSep4 -- also avoid repeating how narrow the data is etc. -- it's good to point out limitations, but repeating them too much is not helpful -- this is the biggest dataset for this question on earth, we need to focus on what we learned from it, not so much on the limitations. Also, Europeans would not call all of the forests northern.

\subsection{Variation across continents and species types}

Our results highlight that spring events show much weaker clines than fall events in tree species, but suggest important variation between species types in spring events. Angiosperms tended to budburst earlier in provenances that were warmer and more southerly, while gymnosperms trended in the opposite direction. Such differences could be driven by their varying investment strategies. As gymnosperms photosynthesize without leafout they generally leafout much later than angiosperms, after most risk of spring frost \citep{panchen14}, and thus may avoid frost risk more than angiosperms. In contrast, angiosperms may tend towards earlier leafout in warmer climates to compete best for access to light and other resources \citep{cat2019}. Testing these hypotheses would require More information on frost risk and forest community assembly from across the provenance locations. 
%EMWSep4 -- I like the below (and I think it may have been my idea), but we'd likely be asked to test for it ... and I am not 100% sure it makes sense -- both angiosperms and conifers need to activate their xylem to photosynthesize I believe. 
% Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring. According to \citet{salk20}, species with narrow xylem such as coniferous gymnosperms and diffuse-porous angiosperms did not exhibit the consistent phenological trends shown in ring-porous species. Therefore, there might be additional differences if we have more species and more information on leafout timing and frost risk approach.

Fall event clines were more similar across species types, but diverged by continent with the common clines of later events in warmer, more southerly latitudes found only in North American studies. However, as we were focused on spring events, we had limited data on fall events, which likely explains our results which contrast with meta-analyses of fall events in Europe (citeAlbero?). In our dataset, more data on fall events were reported in North America, while only two of the six European gardens reported leaf senescence-related metrics, one studying \emph{Fagus sylvatica} and one studying \emph{Fraxinus excelsior}, and neither showed strong clines. In North America studies of both gymnosperm and angiosperm species reported strong clines, and most often focused on budset, while the two European gardens reported data on `leaf senescence' (EXPLAIN what their metric was?, cite Table S1). Budset is most closely related to growth cessation and tied to photoperiod signals, making it generally considered a more robust indicator of when plants stop investing in growth and more genetically controlled than leaf senescence \citep{Alberto13}.
%EMWSep4 -- I am actually not sure if `our results which contrast with meta-analyses of fall events in Europe' is true -- is it? I could not tell from reviewing Alberto very quickly. If it is true from that paper or another, then I think this is the best explanation of our results. If it's not true, I can adjust the text.
% Also, can you say how the two gardens defined `leaf senescence' inside (EXPLAIN)?

%EMWSep4 -- I like to have one paragraph in a paper (usually) focused on biases and limitations, so the below is that paragraph. I also aimed for it to have more of a message specific to this work beyond 'we need more data' (as we always need more data, you can write that in any paper). 
Taken together, these results could explain some of the existing debate on the strength of spring clinal variation, but also highlight how data biases make disentangling the drivers of variation difficult. Angiosperm species showed a weak trend towards earlier budburst in populations from warmer locations. In contrast fall event clines in Europe were weak, but their fall event type (leaf senescence) is less likely to show clinal variation. Given European studies were biased towards studying angiosperms (3/4 species) that are often canopy-species (e.g., \emph{Fagus}, \emph{Quercus}) these results together could explain a greater focus on clinal variation in spring events across European studies (CITESa). Most North American gardens and provenances included in this study were limited to the Pacific Northwest region, and thus tended to focus on species from forests where gymnosperms are almost always the canopy species (\emph{Pseudotsuga, Tsuga, Picea, Pinus}) and angiosperms are much smaller, more opportunistic species (\emph{Alnus, Populus}). This might suggest a far different pattern if gardens included more evenly sampled North American tree species (Fig. \ref{figure:map_gardens}). Given the tremendous investment required for such gardens, however, it is not surprising they are often limited to one species---most often those relevant for forestry (CITES). Our results, however, highlight the limited inference we may gain from such an approach to understand trends across species more broadly. 
%EMWSep4 -- Great point you made about the bias in North American study locations, I built on it. 
% CITESa -- could be Basler & Korner, Gauzere 2020 paper .. you likely have more ideas to add.

\subsection{Adaptation in tree phenology: Implications for climate change responses}

%EMWSep4 -- good stuff here! I tried to refocus on spring events some as our data is focused there. And I limited the amount of hand-wavey stuff we discussed as reviewers only usually want so much extrapolation of your results (so I cut some of the ending stuff). 
Our results suggest that current advances in spring phenology will likely continue, but predicting any shifts due to provenance effects will require new data across more species. Plastic responses to warming, as our results suggest for spring events, mean species can adjust to the shifting environments of climate change---up to some point. Beyond that point theory suggests plastic traits may limit adaptation to climate change (CITE chevin I think?). This may mean species will be replaced when their plastic spring phenological responses are no longer adequate, or shifting reaction norms could lead to adaptation (cite Gauzere 2020?).

Predicting this will require better understanding how different species strategies covary with early-season risks, and how such risks may shift with continued climate change. For temperate zones, many species strategies appear designed to cope with frost, either via avoidance or tolerance \citep{alberto11, lenz16, allevato19}, as utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15}. Climate change at once shifts spring phenology---thus potentially changing each species frost risk---but also appears to reshape the drivers of frost climatically \citep{cat2021pep}. Layered onto this are possible shifts in early-season herbivory with warming \citep{meineke2019}, and trade-offs in early-season risks with access to a long growing season for growth and reproduction, with some species avoiding competition through being very early \citep{guo22}. Common garden studies that track and report frost and herbivory damage, alongside timing of flowering and fruiting, could help to understand the complex fitness landscape of spring phenology.

In contrast to spring events which were weak and variable, fall events exhibited strong clines, which appeared co-gradient with the environment (i.e. late-season events advanced earlier with northward latitudes). This supports previous results and the general theory that fall events are consistently designed to avoid tissue loss at the end of the season, when the cost of such losses could be particularly high \citep{AitkenBemmels16, Alberto13}. Spring events showed no clear trends, but Increasing research into counter-gradient variation for flowering events could be relevant when spring vegetative and flowering events are linked \citep{dan2021nph}.

Understanding environmental clines in phenology will ultimately be critical to predicting how species shift their ranges as well.  Implications of small differences in spring phenology could mean different levels of gene flow, while co- and counter-gradient variation have important implications for gene glow over larger geographical regions (CITES), and thus could shape future range shifts. Much like frost risk, climate change likely shifts multiple factors at once, as climate change shifts pollinators and wind patterns (CITESb). With the anticipated escalation of global temperatures in the forthcoming century, these mechanisms are poised to play a pivotal role in shaping the dynamics of plant communities and the overall carbon balance of ecosystems.
% CITESb Ackerly has a paper on shifting wind patterns in PNAS in the last couple years to cite here.

% All of above are critical to predicting not just phenological change, but range expansions with climate change ... An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.

%EMWSep4 -- I know I wanted this included, but not sure it works. Let's leave out for now. 
%  While some previous chamber studies have found potential adaptation, these could have been influenced by chilling at the provenance site \citep{ettinger20}. However, field studies and common garden experiences do not have this effect. Spring events from the studies we looked at appeared to exhibit high plasticity, while fall events showed strong clines, which is consistent with previous work \citep{AitkenBemmels16, Alberto13}. These findings have important implications for growing discussion of how shifting phenology and tree growth relate as our results suggest there are strong limits on shifts in fall phenology \citep{zohner23, zani20}.

% Our observations reveal that although temperature by itself could serve as a reasonable indicator for predicting when fall events occur across various locations, it's crucial to acknowledge that the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}. High-latitude sites, for example, are likely restricted by the impact of photoperiod \citep{Gill15}. Diverse mechanisms that govern the timing of leaf senescence at the ecosystem level hold the potential to exert significant influence over not only the interactions occurring within plant communities but also the intricate process of carbon storage within ecosystems.
